---
title: "PM 566 HW 4"
author: "Chris Hanson"
date: "11/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```

```{r}
library(dplyr)
```


# HPC

## Problem 1: Make sure your code is nice

**Rewrite the following R functions to make them faster.**

```{r}
mat1 <- matrix(1:80, nrow = 5, ncol = 16, byrow=TRUE)
```

```{r}
# Total row sums
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}
```

```{r}
fun1alt <- function(mat) {
  ans <- rowSums(mat)
  ans
}
```

```{r}
# Cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}
```

```{r}
fun2alt <- function(mat) {
  ans <- t(apply(mat, 1, cumsum))
  ans
}
```

```{r}
# Use the data with this code
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)

# Test for the first
microbenchmark::microbenchmark(
  fun1(dat),
  fun1alt(dat), unit = "relative", check = "equivalent"
)

# Test for the second
microbenchmark::microbenchmark(
  fun2(dat),
  fun2alt(dat), unit = "relative", check = "equivalent"
)
```

## Problem 2: Make things run faster with parallel computing

**The following function allows simulating PI**

(runif generates uniform random numbers between 0 and 1. we generate 2 of these, square them, and add them together. if this is less than 1, we get a 1. if not, we get a 0. we take the mean of 2000 of these and multiply by 4, and this approximates pi, much to my disbelief)

```{r}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

**In order to get accurate estimates, we can run this function multiple times, with the following code:**

(lapply returns a list of the same length as X, each element of which is the result of applying FUN to the corresponding element of X. n is an optional argument to FUN. this runs sim_pi 4000 times, each time with n = 10,000. unlist just turns this from a list to a vector)

```{r}
# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
```

**Rewrite the previous code using parLapply() to make it run faster. Make sure you set the seed using clusterSetRNGStream():**

```{r}
library(parallel)
system.time({
  
  cl <- makePSOCKcluster(5L)
  
  clusterSetRNGStream(cl, 1231)
  
  clusterExport(cl, c("sim_pi"), envir = environment())
  
  ans <- unlist(parLapply(cl = cl, 1:4000, sim_pi, n = 10000))
  print(mean(ans))
  
  stopCluster(cl)
  
  ans
})
```



