---
title: "PM 566 Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chris Hanson
## September 24, 2021
#  
### In this assignment I am given the task of determining whether daily concentration of PM2.5 has decreased in California in the last 15 years. To do this I am using data provided by the U.S. EPA.
<br>
<br>
```{r, message = FALSE}
rm(list = ls())

# Load libraries
library(tidyverse)
library(data.table)
library(janitor)
library(leaflet)
library(ggplot2)

```


### **Problem 1.** _Read in the data. Check dimensions, headers, footers, variable names, and variable types. Check for any data issues, particularly in the key variables. Summarize all findings._
```{r load and check data}
# Read in the data using data.table().
pm2004 <- data.table::fread("ad_viz_plotval_data_2004.csv")
pm2004 <- clean_names(pm2004)
pm2019 <- data.table::fread("ad_viz_plotval_data_2019.csv")
pm2019 <- clean_names(pm2019)
# Check dimensions
paste("The 2004 dataset has", dim(pm2004)[1], "observations and", dim(pm2004)[2], "variables.")
paste("The 2019 dataset has", dim(pm2019)[1], "observations and", dim(pm2019)[2], "variables.")

# Check headers
head(pm2004)
head(pm2019)

# Check footers
tail(pm2004)
tail(pm2019)

# Check variable names
paste(colnames(pm2004))

# Check variable types
sapply(pm2004, typeof)

# Check for issues in key variables:
str(pm2004) # General readout
summary(pm2004) #Summary statistics
summary(pm2004$daily_mean_pm2_5_concentration) #Key variable
mean(is.na(pm2004$daily_mean_pm2_5_concentration)) #No missing values
hist(pm2004$daily_mean_pm2_5_concentration, breaks = 100, xlim=c(0,100)) #Check for outliers

str(pm2019) # General readout
summary(pm2019) #Summary statistics
summary(pm2019$daily_mean_pm2_5_concentration) #Key variable
mean(is.na(pm2019$daily_mean_pm2_5_concentration)) #No missing values
hist(pm2019$daily_mean_pm2_5_concentration, breaks = 100, xlim=c(0,100)) #Check for outliers

```

Summary of findings: 
While the Max values of PM 2.5 in each of these data sets is extremely high, they don't seem to be errors - just extreme events.


### **Problem 2.** _Combine the two years of data into one data frame. Use the Date variable to create a new column for year, which will serve as an identifier. Change the names of the key variables so that they are easier to refer to in your code._

```{r combine and clean data}
# Combine the two years of data into one data frame.
pm <- rbind(pm2004, pm2019)

# Use the Date variable to create a new column for year, which will serve as an identifier.
dates <- pm$date
dates2 <- as.POSIXct(dates, format = "%m/%d/%Y")
dates3 <- format(dates2, format="%Y") 

pm <-  mutate(pm, year = dates3) #New column "year" added

# Change the names of the key variables so that they are easier to refer to in your code.
pm <- rename(pm, pm25 = daily_mean_pm2_5_concentration)

pm$year <- as.numeric(pm$year) #so that colorNumeric works, but can instead use colorFactor

```

### **Problem 3.** _Create a basic map in leaflet() that shows the locations of the sites (make sure to use different colors for each year). Summarize the spatial distribution of the monitoring sites._

```{r}
# Create a basic map in leaflet() that shows the locations of the sites (make sure to use different colors for each year).
pal <- colorNumeric(palette = "RdYlBu", domain=c(2004, 2019)) 
#colorNumeric doesn't work any more

leaflet(pm) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat=~site_latitude,lng=~site_longitude, color=pal(pm$year), opacity=1, fillOpacity=1, radius=100)
  #addLegend('bottomleft', pal=pal, values=~c(2004, 2019),
          #title='Site Locations', opacity=1)
    #this legend is driving me crazy

```

Summarize the spatial distribution of the monitoring sites: ???

### **Problem 4.** _Check for any missing or implausible values of PM2.5 in the combined dataset. Explore the proportions of each and provide a summary of any temporal patterns you see in these observations._

```{r}

```

### **Problem 5.** _Explore the main question of interest at three different spatial levels. Create exploratory plots (e.g. boxplots, histograms, line plots) and summary statistics that best suit each level of data. Be sure to write up explanations of what you observe in these data._
_State_
_County_
_Site in Los Angeles_
