---
title: "PM 566 Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chris Hanson
## September 24, 2021
#  
### In this assignment I am given the task of determining whether daily concentration of PM2.5 has decreased in California in the last 15 years. To do this I am using data provided by the U.S. EPA.
<br>
<br>
```{r, cache = TRUE, results = FALSE}
# Load libraries
library(tidyverse)
library(data.table)

```


**Problem 1.** _Read in the data. Check dimensions, headers, footers, variable names, and variable types. Check for any data issues, particularly in the key variables. Summarize all findings._
```{r load and check data}
# Read in the data using data.table().
pm2004 <- data.table::fread("ad_viz_plotval_data_2004.csv")
pm2019 <- data.table::fread("ad_viz_plotval_data_2019.csv")
# Check dimensions
paste("The 2004 dataset has", dim(pm2004)[1], "observations and", dim(pm2004)[2], "variables.")
paste("The 2019 dataset has", dim(pm2019)[1], "observations and", dim(pm2019)[2], "variables.")

# Check headers
head(pm2004)
head(pm2019)

# Check footers
tail(pm2004)
tail(pm2019)

# Check variable names
paste(colnames(pm2004))

# Check variable types
sapply(pm2004, typeof)

# Check for issues in key variables:
str(pm2004)

```

Summary of findings: 
