---
title: "PM 566 Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chris Hanson
## September 24, 2021
#  
### In this assignment I am given the task of determining whether daily concentration of PM2.5 has decreased in California in the last 15 years. To do this I am using data provided by the U.S. EPA.
<br>
<br>
```{r, results = FALSE}
# Load libraries
library(tidyverse)
library(data.table)
library(janitor)

```


**Problem 1.** _Read in the data. Check dimensions, headers, footers, variable names, and variable types. Check for any data issues, particularly in the key variables. Summarize all findings._
```{r load and check data}
# Read in the data using data.table().
pm2004 <- data.table::fread("ad_viz_plotval_data_2004.csv")
pm2004 <- clean_names(pm2004)
pm2019 <- data.table::fread("ad_viz_plotval_data_2019.csv")
pm2019 <- clean_names(pm2019)
# Check dimensions
paste("The 2004 dataset has", dim(pm2004)[1], "observations and", dim(pm2004)[2], "variables.")
paste("The 2019 dataset has", dim(pm2019)[1], "observations and", dim(pm2019)[2], "variables.")

# Check headers
head(pm2004)
head(pm2019)

# Check footers
tail(pm2004)
tail(pm2019)

# Check variable names
paste(colnames(pm2004))

# Check variable types
sapply(pm2004, typeof)

# Check for issues in key variables:
str(pm2004) # General readout
summary(pm2004) #Summary statistics
summary(pm2004$daily_mean_pm2_5_concentration) #Key variable
mean(is.na(pm2004$daily_mean_pm2_5_concentration)) #No missing values
hist(pm2004$daily_mean_pm2_5_concentration, breaks = 100, xlim=c(0,100)) #Check for outliers

str(pm2019) # General readout
summary(pm2019) #Summary statistics
summary(pm2019$daily_mean_pm2_5_concentration) #Key variable
mean(is.na(pm2019$daily_mean_pm2_5_concentration)) #No missing values
hist(pm2019$daily_mean_pm2_5_concentration, breaks = 100, xlim=c(0,100)) #Check for outliers

```

Summary of findings: 
While the Max values of PM 2.5 in each of these data sets is extremely high, they don't seem to be errors - just extreme events.


**Problem 2.** _Combine the two years of data into one data frame. Use the Date variable to create a new column for year, which will serve as an identifier. Change the names of the key variables so that they are easier to refer to in your code._

```{r combine and clean data}
# Combine the two years of data into one data frame.
pm <- rbind(pm2004, pm2019)

# Use the Date variable to create a new column for year, which will serve as an identifier.
dates <- pm$date
dates2 <- as.POSIXct(dates, format = "%m/%d/%Y")
dates3 <- format(dates2, format="%Y") 

pm <-  mutate(pm, year = dates3) #New column "year" added

# Change the names of the key variables so that they are easier to refer to in your code.

```

