---
title: "PM 566 Week 11 Lab"
author: "Chris Hanson"
date: "11/4/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
```{r setting_up}
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
actor <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/actor.csv")
rental <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/rental.csv")
customer <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/customer.csv")
payment <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/payment_p2007_01.csv")

# Copy data.frames to database
dbWriteTable(con, "actor", actor)
dbWriteTable(con, "rental", rental)
dbWriteTable(con, "customer", customer)
dbWriteTable(con, "payment", payment)

dbListTables(con)
```

# Exercise 1

**Retrieve the actor ID, first name and last name for all actors using the actor table. Sort by last name and then by first name.**

```{r ex1}
dbGetQuery(con,
"SELECT actor_id, first_name, last_name
FROM actor
ORDER BY last_name, first_name
")
```

# Exercise 2

**Retrieve the actor ID, first name, and last name for actors whose last name equals 'WILLIAMS' or 'DAVIS'.**

```{r ex2}
#dbGetQuery(con,
#"SELECT

           
#          ")
```

# Exercise 3

**Write a query against the rental table that returns the IDs of the customers who rented a film on July 5, 2005 (use the rental.rental_date column, and use the date() function to ignore the time component). Include a single row for each distinct customer ID.**

```{r ex3}

```

# Exercise 4

## Exercise 4.1

**Construct a query that retrieves all rows from the payment table where the amount is either 1.99, 7.99, or 9.99.**

```{r ex4-1}

```

## Exercise 4.2

**Construct a query that retrieves all rows from the payment table where the amount is greater then 5.**

```{r ex4-2}

```

# Exercise 5

**Retrieve all payment IDs and their amount from the customers whose last name is ‘DAVIS’.**

```{r ex5}

```

# Exercise 6

## Excercise 6.1

**Use COUNT() to count the number of rows in rental.**

```{r ex6-1}

```

## Excercise 6.2

**Use COUNT() and GROUP BY to count the number of rentals for each customer_id.**

```{r ex6-2}

```

## Excercise 6.3

**Repeat the previous query and sort by the count in descending order.**

```{r ex6-3}

```

## Excercise 6.4

**Repeat the previous query but use HAVING to only keep the groups with 40 or more.**

```{r ex6-4}

```

# Exercise 7

## Excercise 7.1

**Modify the above query to do those calculations for each customer_id.**

```{r ex7-1}

```

## Excercise 7.2

**Modify the above query to only keep the customer_ids that have more then 5 payments.**

```{r ex7-2}

```

# Cleanup

**Run the following chunk to disconnect from the connection.**

```{r cleanup}
# clean up
dbDisconnect(con)
```

