---
title: 'PM 566 Assignment #2'
author: "Chris Hanson"
date: "10/8/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment description: 
Data from USC's Children's Health Study will be used to conduct data wrangling and visualization.
There are two data sets: individual and regional. The individual data includes personal and health characteristics of children in 12 communities across Southern California. The regional data includes air quality measurements at the community level.

```{r downloading-datasets}

```

```{r merging-datasets}

```

## Data wrangling

1. Making sure there are no duplicates by counting the number of rows:
```{r counting-rows}

```

Imputing data to fill missing values:
```{r missing-data}

```

2. Creating variable "obesity_level" and creating a summary table:
```{r categorizing-BMI}

```

3. Creating variable "smoke_gas_exposure":
```{r categorizing-smoke-gas-exposure}

```

4. Creating summary tables for lung expiratory volume vs variables:
```{r expiration-vs-town}

```

```{r expiration-vs-sex}

```

```{r expiration-vs-obesity}

```

```{r expiration-vs-exposure}

```

## Exploratory data analysis


