---
title: 'PM 566 Assignment #2'
author: "Chris Hanson"
date: "10/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(tidytext)
library(ggplot2)
library(tibble)
library(dplyr)
library(dtplyr)
library(data.table)
library(leaflet)
```

## Assignment description: 
Data from USC's Children's Health Study will be used to conduct data wrangling and visualization.
There are two data sets: individual and regional. The individual data includes personal and health characteristics of children in 12 communities across Southern California. The regional data includes air quality measurements at the community level.

Loading the data:

```{r downloading-datasets}
fn  <- "chs_individual.csv"
fn2 <- "chs_regional.csv"

if (!file.exists(fn))
  download.file("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_individual.csv", destfile = fn)

if (!file.exists(fn2))
  download.file("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_regional.csv", destfile = fn2)

chs_individual <-  read.csv(fn)
chs_individual <- as_tibble(chs_individual)

chs_regional <-  read.csv(fn2)
chs_regional <- as_tibble(chs_regional)

```

Merging the datasets:

```{r merging-datasets}
#merge the two datasets
chs <- merge(
  x     = chs_individual,      
  y     = chs_regional, 
    # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE,
  # List of variables to match
  by.x  = "townname",
  by.y  = "townname", 
  )
```

## Data wrangling

### 1. Making sure there are no duplicates by counting the number of rows:
```{r counting-rows}
count(chs_individual, "townname")
count(chs, "townname")
```

Nothing was duplicated during the merging process.

Ensuring that there were no duplicated rows even before the merge:

```{r removing-duplicate-rows, results = FALSE}
nrow(chs %>% distinct())
```

Imputing data to fill missing values:

```{r missing-data}
chs <- chs %>%
  group_by(male, hispanic) %>%
  mutate_all(funs(ifelse(is.na(.), mean(., na.rm = TRUE),.)))
  #mutate_at(vars(male, hispanic), funs(ifelse(is.na(.), mean(., na.rm = TRUE),.)))
```

Some missing values are supposed to be binary: 0 or 1. I filled NA using mean, which assigned non-integer values to some variables. Let's round these to the nearest integer to clean the data.

```{r}
chs <- chs %>%
  mutate(asthma = round(asthma),
         father_asthma = round(father_asthma),
         mother_asthma = round(mother_asthma),
         wheeze = round(wheeze),
         hayfever = round(hayfever),
         allergy  = round(allergy),
         educ_parent = round(educ_parent),
         smoke = round(smoke),
         pets = round(pets),
         gasstove = round(gasstove))
```


### 2. Creating variable "obesity_level" and creating a summary table:
```{r categorizing-BMI, eval = FALSE}
#FAILED CODE, not evaluated
chs[, obesity_level := fifelse(
  bmi < 14, "underweight",
    fifelse(bmi < 22, "normal",
      fifelse(bmi < 24, "overweight", "obese"))
)]
```

```{r categorizing-BMI-2}
chs <- chs %>%
  mutate(obesity_level = cut(bmi,
                         breaks = c(-Inf, 14, 22, 24, Inf),
                         labels = c("underweight", "normal", "overweight", "obese")))
```

```{r obesity_level-table}

#knitr::kable(summary(chs$obesity_level))
summary(chs$bmi)
summary(chs$obesity_level)
```



### 3. Creating variable "smoke_gas_exposure":
```{r categorizing-smoke-gas-exposure}
chs <- chs %>%
  mutate(smoke_gas_exposure = ifelse(smoke == 1 & gasstove == 1, "Smoke & Gas",
                                ifelse(smoke == 0 & gasstove == 1, "Gas",
                                  ifelse(smoke == 1 & gasstove == 0, "Smoke", "None")
  )))
```

### 4. Creating summary tables for lung expiratory volume vs variables:
```{r expiration-vs-town}
#average of fev and proportion of asthma by town
chs %>%
  group_by(townname) %>%
  summarise_at(vars(fev, asthma), list(name = mean))
```

```{r expiration-vs-sex}
chs %>%
  group_by(male) %>%
  summarise_at(vars(fev, asthma), list(name = mean))
```

```{r expiration-vs-obesity}
chs %>%
  group_by(obesity_level) %>%
  summarise_at(vars(fev, asthma), list(name = mean))
```

```{r expiration-vs-exposure}
chs %>%
  group_by(smoke_gas_exposure) %>%
  summarise_at(vars(fev, asthma), list(name = mean))
```

# Exploratory data analysis

### Following the EDA checklist:

```{r}
dim(chs)
```

```{r}
head(chs)
```
```{r}
str(chs)
```
```{r}
summary(chs$bmi)
summary(chs$fev)
```
```{r}
cor(chs$bmi, chs$fev, use="complete")
cor(chs$pm25_mass, chs$fev, use="complete")
cor(chs$smoke, chs$fev, use="complete")
cor(chs$gasstove, chs$fev, use="complete")
```
```{r}
hist(chs$bmi, breaks = 100)
```
```{r}
hist(chs$fev, breaks = 100)
```

A quick visualization of all of the sites in the study:

```{r}
leaflet(chs) %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat=~lat, lng=~lon, opacity=1, fillOpacity=1, radius=100)
```



### Question 1: What is the association between BMI and FEV?



### Question 2: What is the association between smoke and gas exposure and FEV?



### Question 3: What is the association between PM2.5 exposure and FEV?



# Visualization

### 1. Facet plot showing scatterplots with regression lines of BMI and FEV by "townname".

```{r}
chs <- chs %>%
  mutate(has_asthma = as.character(active_asthma))
```

```{r}
ggplot(chs, aes(x=bmi, y=fev)) +
  geom_point(aes(color = has_asthma)) +
  geom_smooth()+
  facet_wrap(~townname) +
  theme(legend.title = element_text(size = 10),
        legend.text = element_text(size = 7))
```


### 2. Stacked histograms of FEV by BMI category and FEV by smoke/gas exposure. 

```{r}
ggplot(chs, aes(fev, fill=obesity_level)) +
  geom_histogram(binwidth = 50)
```

```{r}
ggplot(chs, aes(fev, fill=smoke_gas_exposure)) +
  geom_histogram(binwidth = 50)
```

### 3. Barchart of BMI by smoke/gas exposure.

```{r}
ggplot(chs, aes(x = smoke_gas_exposure, y = bmi)) +
  geom_col()

# Not sure what we want here. What should the y axis be? The mean BMI from each SGE category?
```


### 4. Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category.

```{r}

gfev + 
  geom_point(aes(x = bmi, color = has_asthma)) + 
  geom_smooth(aes(x = bmi)) 

gfev <- ggplot(chs, aes(y = fev))

gfev + 
  geom_boxplot(aes(x = obesity_level)) 

gfev + 
  geom_boxplot(aes(x = obesity_level, color = has_asthma)) 

gfev + 
  geom_point(aes(x = bmi, color = smoke_gas_exposure)) + 
  geom_smooth(aes(x = bmi))

gfev + 
  geom_boxplot(aes(x = smoke_gas_exposure)) 

gfev + 
  geom_boxplot(aes(x = smoke_gas_exposure, color = has_asthma)) + 
  labs(title = 'FEV by SGE')
  
```

### 5. A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities.

```{r}
pal <- colorNumeric(palette = "plasma", domain = chs$pm25_mass)

leaflet(chs) %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat=~lat, lng=~lon, opacity=1, fillOpacity=1, radius=100, color = ~pal(pm25_mass)) %>%
  addLegend("bottomright", pal = pal, values = ~pm25_mass, title = "PM 2.5", opacity = 1)
```


### 6. A visualization to examine whether PM2.5 mass is associated with FEV.


```{r}


gfev + 
  geom_smooth(aes(x = pm25_mass))

gfev + 
  geom_smooth(aes(x = pm25_mass, color = has_asthma))

#gfev + 
  #geom_smooth(aes(x = pm25_mass, color = obesity_level))

gfev + 
  geom_boxplot(aes(x = pm25_mass, group = cut_width(pm25_mass, 7)), width = 5) 

ggplot(chs, aes(x = fev)) + 
  geom_boxplot(aes(y = reorder(townname, -fev, FUN = median), color = pm25_mass))
  
```



