---
title: "PM 566 Lab 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chris Hanson
### September 17, 2021


#### The objective of the lab is to examine the association between weekly average dew point temperature and wind speed in four regions of the US and by elevation.

```{r packages, include = FALSE, cache=TRUE}
library(data.table)
library(tidyverse)
```


```{r, cache=TRUE}
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")
```
## 2. Prepare the data

```{r data-prep}
# Remove temperatures less than -17C
met <- met[temp >= -17]

# Make sure there are no missing data in the key variables coded as 9999, 999, etc
# temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev
met[, range((temp))]
met[, range(rh, na.rm = TRUE)]
met[, range(wind.sp, na.rm = TRUE)]
met[, range(vis.dist, na.rm = TRUE)]
met[, range(dew.point, na.rm = TRUE)]
met[, range(lat, na.rm = TRUE)]
met[, range(lon, na.rm = TRUE)]
met[, range(elev, na.rm = TRUE)]

met[elev == 9999.0, elev :=NA]

# Generate a date variable using the functions as.Date() (hint: You will need the following to create a date paste(year, month, day, sep = "-")).
met[, ymd := as.Date(paste(year, month, day, sep = "-"))]

# Using the data.table::week function, keep the observations of the first week of the month.
met[, table(week(ymd))]
met <- met[ week(ymd) == 31]

# Compute the mean by station of the variables temp, rh, wind.sp, vis.dist, dew.point, lat, lon, and elev.
met_avg = met[, .(
  temp = mean(temp, na.rm = TRUE),
  rh = mean(rh, na.rm = TRUE),
  wind.sp = mean(wind.sp, na.rm = TRUE),
  vis.dist = mean(vis.dist, na.rm = TRUE),
  dew.point = mean(dew.point, na.rm = TRUE),
  lat = mean(lat, na.rm = TRUE),
  lon = mean(lon, na.rm = TRUE),
  elev = mean(elev, na.rm = TRUE), USAFID
), by = "USAFID"]

# Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees
met_avg[lat >= 39.71 & lon <= -98, region := "Northwest"]
met_avg[lat < 39.71 & lon <= -98, region := "Southwest"]
met_avg[lat >= 39.71 & lon > -98, region := "Northeast"]
met_avg[lat < 39.71 & lon > -98, region := "Southeast"]


# Create a categorical variable for elevation as in the lecture slides
met_avg[, elev_cat := fifelse(elev > 252, "high", "low")]

# Deleting extra column
met_avg <- met_avg[, -1]

```
## 3. Use geom_violin to examine the wind speed and dew point temperature by region

This is my attmept at replicating his in class code, but I couldn't copy it in time
```{r}
# Use facets. Make sure to deal with NA category. Describe what you observe in the graph.
ggplot(met_avg, mapping = aes(y = wind.sp, x = 1)) +
  geom_violin() +
  facet_grid(~region)

```
## 4. Use geom_jitter with stat_smooth to examine the association between dew point temperature and wind speed by region.

```{r}
# Color points by region

# Make sure to deal with NA category

# Fit a linear regression line by region

# Describe what you observe in the graph

```


## 5. Use geom_bar to create barplots of the weather stations by elevation category colored by region.

```{r}
# Bars by elevation category using position="dodge"

#Change colors from the default. Color by region using scale_fill_brewer (see note)

# Create nice labels on axes and add a title

# Describe what you observe in the graph

# Make sure to deal with NA

```


## 6. Use stat_summary to examine mean dew point and wind speed by region with standard deviation error bars.

```{r plot-stat_summ}
# Make sure to remove NA. Use fun.data="mean_sdl" in stat_summary. Add another layer of stats_summary but change the geom to "errorbar" (see the help). Describe the graph and what you observe. What is the dew point temperature? What is the wind speed?

ggplot(met_avg[!is.na(wind.sp) &  !is.na(dew.point)]) +
  mapping = aes(x = region y = wind.sp)) +
  stat_summary(fun.data = "mean_sdl") +
  ## UNFINISHED
  
```

## 7. Make a map showing the spatial trend in relative h in the US

```{r leaflet}
# Make sure to remove NA. Use leaflet(). Make a color palette with custom colors. Use addMarkers to include the top 10 places in relative h (hint: rank(-rh) <= 10). Add a legend. Describe trend in RH across the US.

library(leaflet)

temp.pal <- colorNumeric(c('darkgreen', 'goldenrod', 'brown'),
                         domain=)

```

## 8. Use a ggplot extension

```{r}
# Pick an extension (except cowplot) from (link) and make a plot of your choice using the met data (or met_avg). You might watn to try examples that come with the extension first (eg ggtech, gganimate, ggforce). 

```


