---
title: "PM 566 Lab 5"
author: "Chris Hanson"
date: "9/24/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(dtplyr)
library(dplyr)
```

```{r, cache = TRUE}
# Where are we getting the data from
met_url <- "https://github.com/USCbiostats/data-science-data/raw/master/02_met/met_all.gz"
# Downloading the data to a tempfile (so it is destroyed afterwards)
# you can replace this with, for example, your own data:
# tmp <- tempfile(fileext = ".gz")
tmp <- "met.gz"
# We should be downloading this, ONLY IF this was not downloaded already.
# otherwise is just a waste of time.
if (!file.exists(tmp)) {
  download.file(
    url      = met_url,
    destfile = tmp,
    # method   = "libcurl", timeout = 1000 (you may need this option)
  )
}
```


```{r, cache = TRUE}
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")

stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]
```
```{r}
# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]
# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

#removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

#merge the two datasets
met <- merge(
  x     = met,      
  y     = stations, 
    # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE,
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  )

```

## Question 1: What is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the quantile() function. Do these three coincide?

First, generate a representative version of each station. We will use the averages (or median). 

```{r collapsing-by-station}
station_averages <- met[,.(
  temp = mean(temp, na.rm = TRUE),
  wind.sp = mean(wind.sp, na.rm = TRUE),
  atm.press = mean(atm.press, na.rm = TRUE)
), by = USAFID]
```

Now, we need to identify the quantiles per variable.

```{r quantiles}
medians <- station_averages[,.(
  temp_50 = quantile(temp, probs = 0.5, na.rm = TRUE),
  wind.sp_50 = quantile(wind.sp, probs = 0.5, na.rm = TRUE),
  atm.press_50 = quantile(atm.press, probs = 0.5, na.rm = TRUE)
)]

medians
```

Now we can find the stations that are the closest to these. (hint: 'which.min()')

```{r}
station_averages[, temp_dist := abs(temp - medians$temp_50)]
median_temp_station <- station_averages[order(temp_dist)][1]
median_temp_station
```
The station that is closest to the median is `r median_temp_station$USAFID`.

## Question 2: Just like the previous question, you are asked to identify what is the most representative, the median, station per state. This time, instead of looking at one variable at a time, look at the euclidean distance. If multiple stations show in the median, select the one located at the lowest latitude.

```{r}

#We don't have state information in this station_averages dataset, so we'll merge it in from met.
station_averages <- 
  merge(x = station_averages, y=stations, by.x = "USAFID", by.y = "USAF", all.x = TRUE, all.y = FALSE)

#Ok now we can go ahead and...
station_averages[, temp_50 := quantile(temp, probs = 0.5, na.rm = TRUE), by = STATE]
station_averages[, wind.sp_50 := quantile(wind.sp, probs = 0.5, na.rm = TRUE), by = STATE]
station_averages[, atm.press_50 := quantile(atm.press, probs = 0.5, na.rm = TRUE), by = STATE]
```

Now, we compute the euclidean distance: $\sqrt{\sum_i(x_i - y_i)^2}$

```{r}
station_averages[, eudist := sqrt(
  (temp - temp_50)^2 + (wind.sp - wind.sp_50)^2
)]
```


